---
title: "Mussels Plots"
author: "Micaela Chapuis"
date: "2024-07-10"
output: html_document
---

```{r}
library("dplyr")
library("lubridate")
library("stringr")
library("ggplot2")
library("mgcv")
```


### Mussel Plots Data

#### Data Cleaning

Load in Mussel Data

```{r}
mussels <- read.csv("Data/mussel_data.csv")
```

When extracting the bed height data from pictures, if there wasn't a clump of 20 or more mussels at the 10 cm intervals where we took the measurements, then it was marked as NA. But then a lot of data (especially early data) would be excluded from figures and models.

If we didn't have the overlap between A and B plots, we would add up the height of the entire B plot (60 cm) plus the height measurement until a mussel clump in the A plot. But because of the plot overlap (and because it is always different), we decided that the most consistent way of dealing with this is to change the NAs to 60 cm, which is the height of the quadrat. (This makes the data censored because the actual limit of the beds could be outside of the quadrat, that will be dealt with later on)

We have to do it this early on in the script because we should do it for each of the 9 bed height measurements for each quadrat. Once we replace all the NAs, we should calculate the Avg. Height for each plot (each row) again.

```{r}
# goes through the columns with the measurements every 10cm and if there is an NA it will be replaced with the value 60
cols <- c("M1..0", "M2..10", "M3..20", "M4..30", "M5..40", "M6..50", "M7..60", "M8..70", "M9..80")
mussels <- mussels %>% mutate(across(cols, ~ replace(., is.na(.), 60)))
```


```{r}
#take the new average of the values in the individual measurement columns (columns 6 through 14) and store them in Avg.Height
mussels$Avg.Height <- rowMeans(mussels[ , c(6:14)])
```

Get rid of extra columns
```{r}
# since I already calculated the Avg.Height for each plot, I'm getting rid of the columns with the 9 measurements per plot because
# we won't use them
extra.columns <- names(mussels) %in% c("M1..0", "M2..10", "M3..20", "M4..30", "M5..40", "M6..50", "M7..60", "M8..70", "M9..80", "Notes", "Analyzed.By.")

mussels <- mussels[,!extra.columns]
```

Changing column names
```{r}
# Change name of 4th and 5th columns of mussels
names(mussels)[4] <- "Area"
names(mussels)[5] <- "PCover"
```

Make Avg.Height column into an integer
```{r}
mussels$Avg.Height <- as.numeric(mussels$Avg.Height)
```

Dates are weird from the spreadsheet,
Recode dates from Mon-Yr to yr/mm
```{r}
mussels$Date <- dplyr::recode(mussels$Date, "Aug-14" = "2014/08/01", "Aug-16" = "2016/08/01", "Aug-17" = "2017/08/01", "Aug-18" = "2018/08/01", "Dec-14" = "2014/12/01", "Dec-15" = "2015/12/01", "Dec-16" = "2016/12/01", "Dec-17" = "2017/12/01", "Dec-18" = "2018/12/01", "Dec-19" = "2019/12/01", "Jul-14" = "2014/07/01", "Jul-15" = "2015/07/01", "Jul-18" = "2018/07/01", "Jul-19" = "2019/07/01", "Jun-14" = "2014/06/01", "Jun-20" = "2020/06/01", "Mar-15" = "2015/03/01", "Mar-16" = "2016/03/01", "Mar-17" = "2017/03/01", "Mar-18" = "2018/03/01", "Mar-19" = "2019/03/01", "Mar-20" = "2020/03/01", "Mar-21" = "2021/03/01", "Nov-21" = "2021/11/01", "Dec-22" = "2022/12/01", "May-23" = "2023/05/27", "Dec-23" = "2023/12/12", "Mar-24" = "2024/03/07")
```

Turn Date into Date format
(Year-month-day)
```{r}
mussels$Date <- as.Date(mussels$Date, "%Y/%m/%d")
```


Create a Year column
```{r}
mussels$Year <- year(mussels$Date)
```

Add a column for WholePlot (just number, no plot letter, called Location in the paper) + convert WholePlot to numeric
```{r}
mussels[,"WholePlot"] <- NA 
mussels$WholePlot <- str_sub(mussels$Plot, 1, nchar(mussels$Plot)-1) #gets rid of last character, which should be the letter
mussels$WholePlot <- as.numeric(mussels$WholePlot) #was in character, make it numeric
```

Convert Site, Plot, and WholePlot into Factors
```{r}
mussels$Site <- as.factor(mussels$Site)
mussels$Plot <- as.factor(mussels$Plot)
mussels$WholePlot <- as.factor(mussels$WholePlot)
```

Making Date into a continuous variable in a new column (Date.num) and subtracting the start date (June 1st 2014) because otherwise R starts counting from January 1st 1970 (<https://stackoverflow.com/questions/44931645/convert-date-to-continuous-variable-in-r>)
```{r}
mussels <- mussels %>% mutate(Date.num = as.numeric(Date - as.Date("2014-06-01"))) 
plot(mussels$Date, mussels$Date.num) #plotting to make sure it worked
```


#### Data Decisions

Since a lot of the time, the top of the B plot overlapped with the bottom of the A plot, we are taking the average of the A and B plots for our Percent Cover measurements, so from now on we use WholePlot for PCover data.

```{r}
mussels_wholeplot <- mussels %>% group_by(Date, Site, WholePlot, Date.num, Year) %>% summarize(PCover = mean(PCover))
```


Due to the plot overlap, we decided to use data from only B plots for Bed Height measurements, so making a column that just has the plot letter and making a df with just the B plots. But also Plots 4 and 8 don't have B plots, so 4A and 8A need to be included in this new df as well.

```{r}
#subsets out the last character in the "Plot" string, which should always be the plot letter (A or B) and stores it in new column
mussels$Letter <-  str_sub(mussels$Plot, -1)

# only keeping the B plots and plots 4A and 8A
mussels_B <- mussels %>% filter((Letter %in% "B") | (WholePlot == 4) | (WholePlot == 8))
```


## Raw Data Figures

```{r}
group.colors <- c(West = "#ff6361", East = "#619eff")
```


###### Fig S1A - Mussel Cover Raw Data
```{r}
FigS1A <- mussels_wholeplot %>% 
  ggplot(mapping=aes(x = Date, y = PCover, color = Site)) + 
      labs (x = "Date", y = "Mussel Percent Cover") + 
      theme(panel.background = element_blank(), 
          axis.line = element_line (colour = "black"), 
          axis.text = element_text(size = 32),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_text(size = 36),
          legend.position = "none") + 
      geom_point(size=4,alpha = 0.7) + 
      scale_x_date(limits = as.Date(c("2014-06-01", "2024-04-01"))) +
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors)

FigS1A

#ggsave("figS1A.png", height=3, width=3,scale = 3)
```


######Fig S1B - Mussel Bed Height Raw Data

```{r}
FigS1B <- mussels_B %>% 
  ggplot(mapping=aes(x = Date, y = Avg.Height, color = Site)) + 
      labs (x = "Date", y = "Average Mussel Bed Height (cm)") + 
      theme(panel.background = element_blank(), 
          axis.line = element_line (colour = "black"), 
          axis.text = element_text(size = 32),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_text(size = 36),
          legend.position = "none") + 
      geom_point (size = 4, alpha = 0.7) + 
    scale_x_date(limits = as.Date(c("2014-06-01", "2024-04-01"))) +
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors)

FigS1B

#ggsave("figS1B.png", height=3, width=3,scale = 3)
```


### Models

First, we're averaging mussel cover and mussel bed height data by Date and Site to deal with censoring in the bed height data (see paper for details)
```{r}
mussel_cover_avg <- mussels %>% group_by(Date, Site) %>% summarize(PCover = mean(PCover))
mussel_height_avg <- mussels_B %>% group_by(Date, Site) %>% summarize(Avg.Height = mean(Avg.Height))
```

#### Mussel Cover Model 
Converting date to decimal date
```{r}
mussel_cover_avg <- mussel_cover_avg %>% mutate(date_dec = decimal_date(Date))
mussel_cover_avg
```

GAMS
Reference
http://r.qcbs.ca/workshop08/book-en/introduction-to-gams.html

Most of the code from here on was written by Robin Elahi

```{r}
#practice with one subset of the data first
mussel_cover_avg_sub <- mussel_cover_avg %>% filter(Site == "West")
```

```{r}
## Linear
cover.gam1 <- gam(PCover ~ date_dec, data = mussel_cover_avg_sub)
summary(cover.gam1)

par(mfrow = c(2, 2))
gam.check(cover.gam1) # residuals exhibit curvilinear trend
fitted(cover.gam1)
acf(residuals(cover.gam1)) 
```

```{r}
## Wiggly
cover.gam2 <- gam(PCover ~ s(date_dec), data = mussel_cover_avg_sub) #the s() function makes it a smoothed (non-linear) term
summary(cover.gam2) #R squared increased a lot
plot(cover.gam2) # "The mgcv package also includes a default plot() function to look at the smooths"

acf(residuals(cover.gam2)) 
par(mfrow = c(2,2))
gam.check(cover.gam2)
k.check(cover.gam2)
```

```{r}
## Compare models - "Generally, the smaller the AIC, the “better” is the predictive performance of the model."
# Here we ask whether adding a smooth function to the linear model improves the fit of the model to our dataset.

AIC(cover.gam1, cover.gam2) # lower AIC = better performing model --> gam2 (with smoothed term) is better

#"Here, the AIC of the smooth GAM is lower, which indicates that adding a smoothing function improves model performance. Linearity is therefore not supported by our data."
```

```{r} 
#"when we compare the fit of the linear (red) and non-linear (blue) models, it is clear that the blue one is more appropriate for our dataset"
mussel_cover_avg_sub %>% 
  ggplot(aes(date_dec, PCover)) + 
  geom_point(size = 3, alpha = 0.5) + 
  geom_line(aes(y = fitted(cover.gam1)), color = "red") + 
  geom_line(aes(y = fitted(cover.gam2)), color = "blue")
```

Now for real with all the data
```{r}
mussel_cover_avg <- mussel_cover_avg %>% mutate(Site = as.factor(Site))
mussel_cover_avg
```

```{r}
## Trying with no interaction between date and site
cover.gam1 <- gam(PCover ~ s(date_dec) + Site, data = mussel_cover_avg, method = "ML")
gam1_summary <- summary(cover.gam1)
gam1_summary$p.table
gam1_summary$s.table
par(mfrow = c(1, 2))
plot(cover.gam1, all.terms = TRUE)
acf(residuals(cover.gam1)) 
```


```{r}
# USING THIS MODEL (see AIC below)
## With interaction between date and site
# https://r.qcbs.ca/workshop08/book-en/gam-with-interaction-terms.html

cover.gam2 <- gam(PCover ~ s(date_dec, by = Site) + Site, data = mussel_cover_avg, method = "ML") # this is default, thin plate regression spline
gam2_summary <- summary(cover.gam2)
gam2_summary$p.table
gam2_summary$s.table
par(mfrow = c(2, 2))
plot(cover.gam2, all.terms = TRUE) # the first two plots show the interaction effect of the date smooth and each level of the Site factor variable.
acf(residuals(cover.gam2)) 
```

```{r}
## Check gam
par(mfrow = c(2, 2))
gam.check(cover.gam2)
```

```{r}
## Compare models
AIC(cover.gam1, cover.gam2) #gam 1 has no interaction, gam 2 has interaction. This shows that including the interaction between date and site improves the model's performance

#overall, we're keeping GAM over LM, with the smoothed term (date) and with the date*site interaction (cover.gam2)
```

##### Plot with Model Data
```{r}
x_range <- range(mussel_cover_avg$date_dec)
x_new <- seq(from = x_range[1], to = x_range[2], by = 0.1)
x_new

site_levels <- unique(mussel_cover_avg$Site)
```

```{r}
# Make the prediction dataframe
cover_pred <- expand.grid(x_new, site_levels) %>% 
  as_tibble() %>% 
  rename(date_dec = Var1, Site = Var2)
head(cover_pred)
```

```{r}
# Predictions given the model
y_pred <- predict(cover.gam2, cover_pred, se.fit = TRUE)
y_pred
```

```{r}
# 95% CI assuming normal distribution
upr <- y_pred$fit + (2 * y_pred$se.fit) 
lwr <- y_pred$fit - (2 * y_pred$se.fit)

cover_pred <- cover_pred %>%
  mutate(PCover = y_pred$fit, 
         upper = upr, 
         lower = lwr)
```

###### Fig2A - Mussel Cover Model Data
```{r}
## Final plot with model output
Fig2A <- mussel_cover_avg %>% 
  ggplot(aes(date_dec, PCover, color = Site, fill = Site)) + 
  geom_ribbon(data = cover_pred, aes(ymin = lower, ymax = upper, 
                                 fill = Site, color = NULL), alpha = 0.5) + 
  geom_line(data = cover_pred, size =0.75) + 
  geom_point(size = 5, alpha = 0.7, pch = 21, color = "black") +
  theme_minimal() + 
  theme(panel.background = element_blank(), 
        axis.line = element_line (colour = "black"), 
        axis.text.y = element_text(size = 25),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size=25),
        axis.title.y = element_text(size = 25),
        axis.title.x = element_text(size = 25),
        legend.position = "none") +
  scale_y_continuous(limits = c(10, 100)) + 
  labs(x = "Date", y = "Mussel Percent Cover") +
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors) 

Fig2A
#ggsave("Fig2A.png", height=3, width=3, scale=3)
```


#### Mussel Bed Height
Converting date to decimal date
```{r}
mussel_height_avg <- mussel_height_avg %>% mutate(date_dec = decimal_date(Date))
mussel_height_avg
```

GAMS
Reference
http://r.qcbs.ca/workshop08/book-en/introduction-to-gams.html

Most of the code from here on was written by Robin Elahi

```{r}
#practice with one subset of the data first
mussel_height_avg_sub <- mussel_height_avg %>% filter(Site == "West")
```

```{r}
## Linear
height.gam1 <- gam(Avg.Height ~ date_dec, data = mussel_height_avg_sub)
summary(height.gam1)

par(mfrow = c(2, 2))
gam.check(height.gam1) # residuals exhibit curvilinear trend
fitted(height.gam1)
acf(residuals(height.gam1)) 
```


```{r}
## Wiggly
height.gam2 <- gam(Avg.Height ~ s(date_dec), data = mussel_height_avg_sub) #the s() function makes it a smoothed (non-linear) term
summary(height.gam2) #R squared increased a lot
plot(height.gam2) # "The mgcv package also includes a default plot() function to look at the smooths"

acf(residuals(height.gam2)) 
par(mfrow = c(2,2))
gam.check(height.gam2)
k.check(height.gam2)
```

```{r}
## Compare models - "Generally, the smaller the AIC, the “better” is the predictive performance of the model."
# Here we ask whether adding a smooth function to the linear model improves the fit of the model to our dataset.

AIC(height.gam1, height.gam2) # lower AIC = better performing model --> gam2 (with smoothed term) is better

#"Here, the AIC of the smooth GAM is lower, which indicates that adding a smoothing function improves model performance. Linearity is therefore not supported by our data."
```

```{r} 
#"when we compare the fit of the linear (red) and non-linear (blue) models, it is clear that the blue one is more appropriate for our dataset"
mussel_height_avg_sub %>% 
  ggplot(aes(date_dec, Avg.Height)) + 
  geom_point(size = 3, alpha = 0.5) + 
  geom_line(aes(y = fitted(height.gam1)), color = "red") + 
  geom_line(aes(y = fitted(height.gam2)), color = "blue")
```

Now for real with all the data
```{r}
mussel_height_avg <- mussel_height_avg %>% mutate(Site = as.factor(Site))
mussel_height_avg
```

```{r}
## Trying with no interaction between date and site
height.gam1 <- gam(Avg.Height ~ s(date_dec) + Site, data = mussel_height_avg, method = "ML")
gam1_summary <- summary(height.gam1)
gam1_summary$p.table
gam1_summary$s.table
par(mfrow = c(1, 2))
plot(height.gam1, all.terms = TRUE)
acf(residuals(height.gam1)) 
```


```{r}
# USING THIS MODEL (see AIC below)
## With interaction between date and site
# https://r.qcbs.ca/workshop08/book-en/gam-with-interaction-terms.html

height.gam2 <- gam(Avg.Height ~ s(date_dec, by = Site) + Site, data = mussel_height_avg, method = "ML") # this is default, thin plate regression spline
gam2_summary <- summary(height.gam2)
gam2_summary$p.table
gam2_summary$s.table
par(mfrow = c(2, 2))
plot(height.gam2, all.terms = TRUE) # the first two plots show the interaction effect of the date smooth and each level of the Site factor variable.
acf(residuals(height.gam2)) 
```

```{r}
## Check gam
par(mfrow = c(2, 2))
gam.check(height.gam2)
```

```{r}
## Compare models
AIC(height.gam1, height.gam2) #gam 1 has no interaction, gam 2 has interaction. This shows that including the interaction between date and site improves the model's performance

#overall, we're keeping GAM over LM, with the smoothed term (date) and with the date*site interaction (height.gam2)
```

##### Plot with Model Data
```{r}
height.x_range <- range(mussel_height_avg$date_dec)
height.x_new <- seq(from = height.x_range[1], to = height.x_range[2], by = 0.1)
height.x_new

height.site_levels <- unique(mussel_height_avg$Site)
```

```{r}
# Make the prediction dataframe
height_pred <- expand.grid(height.x_new, height.site_levels) %>% 
  as_tibble() %>% 
  rename(date_dec = Var1, Site = Var2)
head(height_pred)
```

```{r}
# Predictions given the model
height.y_pred <- predict(height.gam2, height_pred, se.fit = TRUE)
height.y_pred
```

```{r}
# 95% CI assuming normal distribution
height.upr <- height.y_pred$fit + (2 * height.y_pred$se.fit) 
height.lwr <- height.y_pred$fit - (2 * height.y_pred$se.fit)

height_pred <- height_pred %>%
  mutate(Avg.Height = height.y_pred$fit, 
         upper = height.upr, 
         lower = height.lwr)
```

###### Fig2B - Mussel Bed Height Model Data
```{r}
## Final plot with model output
Fig2B <- mussel_height_avg %>% 
  ggplot(aes(date_dec, Avg.Height, color = Site, fill = Site)) + 
  geom_ribbon(data = height_pred, aes(ymin = height.lwr, ymax = height.upr, 
                                 fill = Site, color = NULL), alpha = 0.5) + 
  geom_line(data = height_pred) + 
  geom_point(size = 5, alpha = 0.7, pch = 21, color = "black") +
  theme_minimal() + 
  theme(panel.background = element_blank(), 
        axis.line = element_line (colour = "black"), 
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.text = element_text(size = 25),
        axis.title = element_text(size = 25),
        legend.position="none") +
  labs(x = "Date", y = "Mussel Bed Height (cm)") +  
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors)

Fig2B

#ggsave("Fig2B.png", height=3, width=3, scale=3)

```



## Creating new csv files
For predator models
```{r}
mussel_cover_yearly_avg <- mussels %>% group_by(Year, Site) %>% summarize(PCover = mean(PCover))
mussel_height_yearly_avg <- mussels_B %>% group_by(Year, Site) %>% summarize(Avg.Height = mean(Avg.Height))
#write.csv(mussel_cover_yearly_avg, file = "mussel_cover_yearly_avg.csv", row.names = FALSE)  #to use in models script
#write.csv(mussel_height_yearly_avg, file = "mussel_height_yearly_avg.csv", row.names= FALSE) #to use in models script
```

For abiotic models
```{r}
#write.csv(mussels, file = "mussels.csv", row.names = FALSE)  #to use in models script
#write.csv(mussels_B, file = "mussels_B.csv", row.names = FALSE)  #to use in models script
```


