---
title: "NOAA_Wave_Data"
author: "Micaela Chapuis"
date: "2024-07-09"
output: html_document
---

```{r}
library("dplyr")
library("lubridate")                                 
library("readr")
```

## Making Dataset
Import all NOAA CSV files and join them into one data set
```{r}
noaa_data <- list.files(path = "Data/NOAA_Buoy_46240/Raw NOAA Data/csv files",  # Identify all CSV files in the folder 
  pattern = "*.csv", full.names = TRUE) %>% 
  lapply(read_csv) %>%                              # Store all files in list
  bind_rows                                         # Combine data sets into one data set 

noaa_data                                            # Print data to RStudio console
```


Filter out the first row that includes the units for each column
```{r}
noaa_data <- noaa_data %>% filter(!(noaa_data$MM %in% "mo"))
```


Rename the year, month, and day columns (ignoring times)
```{r}
names(noaa_data)[1] <- "Year"
names(noaa_data)[2] <- "Month"
names(noaa_data)[3] <- "Day"
```


Every single entry is a number
So take all the entries and if they are characters make them numeric
```{r}
noaa_data <- noaa_data %>% mutate_if(is.character,as.numeric)
```


This dataset uses "99" or "999" as a replacement for NA --> change back to NA (only doing it on the columns I'm using for now, others still have the 999)
```{r}
noaa_data$WVHT <- na_if(noaa_data$WVHT, 99)
noaa_data$WVHT <- na_if(noaa_data$WVHT, 999) 
noaa_data$WTMP <- na_if(noaa_data$WTMP, 99)
noaa_data$WTMP <- na_if(noaa_data$WTMP, 999)  
```


Individual components of date are spread across multiple columns, fix by combining it into one column called Date: 
```{r}
noaa_data_dt <- noaa_data %>% select(Year, Month, Day, WVHT, ATMP, WTMP) %>% mutate(Date = make_date(Year, Month, Day))
```


Take the daily average for wave height and water temp
```{r}
daily_waveheight <- noaa_data_dt %>% subset(!(is.na(WVHT))) %>%
  group_by(Date) %>%
  summarize(WVHT= mean(WVHT))
daily_watertemp <- noaa_data_dt %>% subset(!(is.na(WTMP))) %>% 
  group_by(Date) %>%
  summarize(WTMP= mean(WTMP))
```


Join them
```{r}
daily_wavedata <- full_join(daily_waveheight, daily_watertemp, by = "Date")
```


Save new dataframe as csv file
```{r}
write.csv(daily_wavedata, file = "daily_wavedata.csv", row.names = FALSE)
```

## Using Data
```{r}
daily_wavedata$Date <- as.Date(daily_wavedata$Date)
daily_wavedata$Year <- year(daily_wavedata$Date)
```

Rename WVHT and WTMP
```{r}
daily_wavedata <- daily_wavedata %>% rename("Wave.Height" = "WVHT",
                                            "Water.Temp" = "WTMP")
```

Plot WVHT and WTMP over time
```{r}
plot(daily_wavedata$Date, daily_wavedata$Wave.Height)

plot(daily_wavedata$Date, daily_wavedata$Water.Temp)
```
### Raw Data Figures

##### Fig S1E - Wave Height Raw Data
Daily Average Wave Height over Time
```{r}
FigS1E <- daily_wavedata %>% 
  subset(Year > 2013) %>%
  ggplot(mapping=aes(x = Date, y = Wave.Height)) + 
      labs (x = "Date", y = "Wave Height (m)") +  
      theme(panel.background = element_blank(), 
          axis.line = element_line (colour = "black"), 
          axis.text = element_text(size = 32),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_text(size = 36)) + 
      geom_point (size=3,alpha = 0.7)+ 
      theme(plot.title = element_text(hjust = 0.5, size=35, face="bold"))

FigS1E
#ggsave("figS1E.png", height=3, width=3,scale = 3)
```
##### Fig S1F - Water Temperature Raw Data
Daily Average Water Temperature over Time

```{r}
FigS1F <- daily_wavedata %>% 
  subset(Year > 2013) %>%
  ggplot(mapping=aes(x = Date, y = Water.Temp)) + 
      labs (x = "Date", y = "Water Temperature (°C)") + 
      theme(panel.background = element_blank(), 
          axis.line = element_line (colour = "black"), 
          axis.text = element_text(size = 32),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_text(size = 36)) + 
      geom_point (size=3, alpha = 0.7) + 
      scale_y_continuous(breaks = c(10, 13, 16, 19)) +
      theme(plot.title = element_text(hjust = 0.5, size=30, face="bold"))

FigS1F
#ggsave("figS1F.png", height=3, width=3,scale = 3)
```

## Models

Importing the csv made in the Models script, where the wave data was averaged for each of the 3 months preceding each mussel data collection date
```{r}
waves_avg <- read.csv("Data/Data for Models/mussels.abiotic.csv")
```

Turn Date into Date format
(Year-month-day)
```{r}
waves_avg$Date <- as.Date(waves_avg$Date)
```


```{r}
waves_avg <- waves_avg %>% mutate(date_dec = decimal_date(Date))
```

#### Wave Height
```{r}
# THIS IS THE MODEL WE'RE USING
wht.m1 <- lm(Wave.Height ~ date_dec, data = waves_avg)
summary(wht.m1)
acf(residuals(wht.m1)) 
tidy(wht.m1)
```
```{r}
whtgam1 <- gam(Wave.Height ~ date_dec, data = waves_avg, method = "ML")
whtgam1_summary <- summary(whtgam1)
whtgam1_summary$p.table
whtgam1_summary$s.table
par(mfrow = c(1, 2))
plot(whtgam1, all.terms = TRUE)
acf(residuals(whtgam1)) 
```

```{r}
AIC(wht.m1, whtgam1) # lower AIC = better performing model --> equal
#so we're keeping the linear model
```
##### Plot with Model Data
######Fig2E - Wave Height Model Data
```{r}
Fig2E <- waves_avg %>% 
  ggplot(aes(date_dec, Wave.Height)) + 
  geom_smooth(method = "lm", fill = "gray79", color = "gray25") + 
  geom_point(size = 5, alpha = 0.7, pch = 21, fill = "black") +
  theme_minimal() +
  theme(panel.background = element_blank(), 
        axis.line = element_line (colour = "black"),
        axis.text.y = element_text(size =25),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 25),
        axis.text = element_text(size = 25),
        axis.title = element_text(size = 25)) + 
  labs(x = "Date", y = "Wave Height (m)") 

Fig2E
#ggsave("Fig2E.png", height=3, width=3, scale=3)
```


#### Water Temperature

```{r}
# THIS IS THE MODEL WE'RE USING
wtmp.m1 <- lm(Water.Temp ~ date_dec, data = waves_avg)
summary(wtmp.m1)
acf(residuals(wtmp.m1)) 
tidy(wtmp.m1)
```

```{r}
wtmpgam1 <- gam(Water.Temp ~ date_dec, data = waves_avg, method = "ML")
wtmpgam1_summary <- summary(wtmpgam1)
wtmpgam1_summary$p.table
wtmpgam1_summary$s.table
par(mfrow = c(1, 2))
plot(wtmpgam1, all.terms = TRUE)
acf(residuals(wtmpgam1)) 
```

```{r}
AIC(wtmp.m1, wtmpgam1) # lower AIC = better performing model --> equal
# so we're using the linear model
```
##### Plot with Model Data
###### Fig 2F - Water Temperature Model Data
```{r}
Fig2F <- waves_avg %>% 
  ggplot(aes(date_dec, Water.Temp)) + 
  geom_smooth(method = "lm", fill = "gray79", color = "gray25") + 
  geom_point(size = 5, alpha = 0.7, pch = 21, fill = "black") +
  theme_minimal() +
  theme(panel.background = element_blank(), 
        axis.line = element_line (colour = "black"), 
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.text = element_text(size = 25),
        axis.title = element_text(size = 25)) + 
  labs(x = "Date", y = "Water Temperature (°C)") 

Fig2F
#ggsave("Fig2F.png", height=3, width=3, scale=3)
```


