---
title: "Predators_Data"
author: "Micaela Chapuis"
date: "2024-07-12"
output: html_document
---

```{r}
library("dplyr")
library("lubridate")
library("ggplot2")
```

## Whelks

```{r}
whelk <- read.csv("Data/whelk_data.csv")
```

### Data Cleaning
Recode "28 + whelk eggs" data point to just "28" and switch Number to numeric

```{r}
whelk$Number <- dplyr::recode(whelk$Number, "28 + whelk eggs" = "28")
whelk$Number <- as.numeric(whelk$Number)
```

Recode Date column + turn into Date format (changing 15-Jul-14 to 14-Jul-14 to combine those)

```{r}
whelk$Date <- dplyr::recode(whelk$Date, "6/18/2014" = "2014-06-18", "6/19/2014" = "2014-06-19", "6/27/2014" = "2014-06-27", "7/15/2014" = "2014-07-14", "7/14/2014" = "2014-07-14", "8/11/2014" = "2014-08-11", "10/1/2015" = "2015-10-01", "7/14/2017" = "2017-07-14", "5/17/2021" = "2021-05-17", "7/7/2023" = "2023-07-07", "12/12/2023" = "2023-12-12", "3/8/2024" = "2024-03-08")
whelk$Date <- as.Date(whelk$Date)
```

Fix name of "Plot" column

```{r}
names(whelk)[3] <- "Plot"
```

Change name of "Number" column to be more descriptive

```{r}
names(whelk)[4] <- "Num.Whelks"
```

Filter out B plots and making new df "whelks" without them (Whelk quadrats, which are smaller than mussel quadrats, were supposed to be put in between the A and B plots, right on their shared border. At some point, people collected whelk data for the A and the B plots separately. We chose to only keep the A plots since not all plots have B plots)
```{r}
B.plots <- c("1B", "2B", "3B", "5B", "6B", "7B", "9B", "10B")
whelks <- whelk %>% filter(!Plot %in% B.plots)
```

Recode A plots to WholePlot numbers

```{r}
whelks$Plot <- dplyr::recode(whelks$Plot, "1A" = "1", "2A" = "2", "3A" = "3", "4A" = "4", "5A" = "5", "6A" = "6", "7A" = "7", "8A" = "8", "9A" = "9", "10A" = "10")
```

Change Date to date and Site and Plot to factor

```{r}
whelks$Date <- as.Date(whelks$Date, "%Y-%m-%d")
whelks$Site <- as.factor(whelks$Site)
whelks$Plot <- as.factor(whelks$Plot)
```

Create column for just year

```{r}
whelks$Year <- year(whelks$Date)
```


### Raw Data Figure
```{r}
group.colors <- c(West = "#ff6361", East = "#619eff")
```

###### Fig S1D - Whelk Abundance Raw Data
```{r}
FigS1D <- whelks %>% 
  ggplot(mapping=aes(x = Date, y = Num.Whelks, color = Site)) + 
      labs (x = "Date", y = "Whelk Abundance (# whelks/0.25m2") + 
      theme(panel.background = element_blank(), 
          axis.line = element_line (colour = "black"), 
          axis.text = element_text(size = 32),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_text(size = 36),
          legend.position = "none") + 
      geom_point (size=4, alpha = 0.7) + 
      scale_x_date(limits = as.Date(c("2014-01-01", "2024-04-01"))) +
    theme(plot.title = element_text(hjust = 0.5, size=30, face="bold"))+
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors) 

FigS1D
#ggsave("figS1D.png", height=3, width=3,scale = 3)

```




## Seastars

```{r}
seastars <- read.csv("Data/seastar_data.csv")
```

Changing name of Count column to Num.Seastars

```{r}
names(seastars)[3] <- "Num.Seastars"
```

### Raw Data Figure
#####FigS1C - Seastar Abundace Raw Data (2014-2024)

```{r}
FigS1C <- seastars %>% 
  subset(Year>2013) %>%
  ggplot(mapping=aes(x = Year, y = Num.Seastars, color = Site)) + 
  labs (x = "Date", y = "Sea Star Abundance (# sea stars/site)") + 
  theme(panel.background = element_blank(), 
        axis.line = element_line (colour = "black"), 
        axis.text = element_text(size = 32),
              axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.title = element_text(size = 36),
        legend.text = element_text(size = 32),
        legend.title = element_text(size = 36)) + 
  geom_point (size=4, alpha = 0.7, position="jitter") + 
  theme(plot.title = element_text(hjust = 0.5, size=35, face="bold")) +
  scale_x_continuous(breaks = c(2014, 2016, 2018, 2020, 2022, 2024)) +
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors)

FigS1C

#ggsave("figS1C.png", height=3, width=3,scale = 3)

```



## Models

#### Seastars

```{r}
sd <- seastars %>% filter(Year > 2013)
#write.csv(sd, file = "seastars_avg.csv", row.names = FALSE)  #to be used in the models script
```

Most of the code from here on (within this section) was written by Robin Elahi

Following this: https://rpubs.com/markpayne/164550
```{r}
#with Year * Site interaction
stars.m1 <- lm(Num.Seastars ~ Year * Site, data = sd)
summary(stars.m1)
acf(residuals(stars.m1)) 
```
```{r}
# THIS IS THE MODEL WE CHOSE
# without Year * Site interaction
stars.m2 <- lm(Num.Seastars ~ Year + Site, data = sd)
summary(stars.m2)
acf(residuals(stars.m2)) 
tidy(stars.m2)
```


```{r}
glimpse(sd)
sd <- sd %>% 
  mutate(Site = as.factor(Site))
sd
```

GAMS
```{r}
## No interaction between date and site
sgam1 <- gam(Num.Seastars ~ Year + Site, data = sd, method = "ML") # won't run with smoothed year
sgam1_summary <- summary(sgam1)
sgam1_summary$p.table
sgam1_summary$s.table
par(mfrow = c(1, 2))
plot(sgam1, all.terms = TRUE)
acf(residuals(sgam1)) 
```

```{r}
## Compare models - "Generally, the smaller the AIC, the “better” is the predictive performance of the model."
# Here we compare the linear model to the GAM

AIC(stars.m2, sgam1) # lower AIC = better performing model --> they're the same
#so sticking with the linear model
```

##### Plot with Model Data
######Fig2C - Seastars Model Data
```{r}
Fig2C <- sd %>% 
  ggplot(aes(Year, Num.Seastars, color = Site, fill = Site)) + 
  geom_smooth(method = "lm", aes(group=NULL), fill = "gray79", color = "gray25") + 
  geom_point(size = 5, alpha = 0.7, pch = 21, color = "black") +
  theme_minimal() + 
  theme(panel.background = element_blank(), 
        axis.line = element_line (colour = "black"), 
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.text = element_text(size = 25),
        axis.title = element_text(size = 25),
        legend.position = "none") +
  labs(x = "Year", y = "Sea Star Abundance") +
  scale_x_continuous(breaks = c(2014, 2016, 2018, 2020, 2022, 2024)) +
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors) 

Fig2C

#ggsave("Fig2C.png", height=3, width=3, scale=3)
```

####Long Term Seastars
```{r}
lt_stars <- seastars %>% group_by(Year,Site) %>% summarize(Num.Seastars = mean(Num.Seastars))
```

Most of the code from here on (within this section) was written by Robin Elahi

Following this: https://rpubs.com/markpayne/164550
```{r}
#linear model with year site interaction
lts.m1 <- lm(Num.Seastars ~ Year * Site, data = lt_stars)
summary(lts.m1)
acf(residuals(lts.m1)) 
```

```{r}
# THIS IS THE MODEL WE CHOSE
#linear model without year site interaction
lts.m2 <- lm(Num.Seastars ~ Year + Site, data = lt_stars)
summary(lts.m2)
acf(residuals(lts.m2)) 
```
```{r}
glimpse(lt_stars)
lt_stars <- lt_stars %>% 
  mutate(Site = as.factor(Site))
lt_stars
```

```{r}
## No interaction between date and site
ltsgam1 <- gam(Num.Seastars ~ Year + Site, data = lt_stars, method = "ML") # won't run with smoothed year
ltsgam1_summary <- summary(ltsgam1)
ltsgam1_summary$p.table
ltsgam1_summary$s.table
par(mfrow = c(1, 2))
acf(residuals(ltsgam1)) 
```


```{r}
## Compare models - "Generally, the smaller the AIC, the “better” is the predictive performance of the model."
# Here we compare linear model to GAM

AIC(lts.m2, ltsgam1) # lower AIC = better performing model --> they're the same
#so we're using the linear model
```
##### Plot with Model Data
######Fig 3 - Long Term Seastar Abundance
```{r}
Fig3 <- lt_stars %>% 
  ggplot(aes(Year, Num.Seastars, color = Site, fill = Site)) + 
  geom_smooth(method = "lm", aes(group=NULL), fill = "gray79", color = "gray25") + 
  geom_point(size = 5, alpha = 0.7, pch = 21, color = "black") +
  theme_minimal() + 
  theme(panel.background = element_rect(color="white"),
        axis.line = element_line (colour = "black"), 
        axis.text.x = element_text(angle = 30, hjust = 0.8),
        axis.text = element_text(size = 25),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 25),
        legend.background = element_rect(color=NA)) +
  labs(x = "Year", y = "Sea Star Abundance") +
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors) +
  scale_x_discrete(limits = c(1954, 1970, 1985, 1995, 2004, 2014, 2024))

Fig3

#ggsave("Fig3.pdf", height=3, width=4, scale=2.5, bg="white")
```


#### Whelks

Averaging Whelks to Site and Year to match Seastars
```{r}
wd <- whelks %>% group_by(Site, Year) %>% summarize(Num.Whelks = mean(Num.Whelks))
#write.csv(wd, file = "whelks_avg.csv", row.names = FALSE)  #to use in models script
```


Most of the code from here on was written by Robin Elahi
Following this: https://rpubs.com/markpayne/164550
```{r}
#with Year*Site interaction
whelks.m1 <- lm(Num.Whelks ~ Year * Site, data = wd)
summary(whelks.m1)
acf(residuals(whelks.m1)) 
```


```{r}
#THIS IS THE MODEL WE CHOSE
#without Year*Site interaction
whelks.m2 <- lm(Num.Whelks ~ Year + Site, data = wd)
summary(whelks.m2)
acf(residuals(whelks.m2)) 
tidy(whelks.m2)
```


```{r}
AIC(whelks.m1, whelks.m2) #without the interaction = better
```

```{r}
glimpse(wd)
wd <- wd %>% 
  mutate(Site = as.factor(Site))
wd
```

```{r}
## No interaction between date and site
wgam1 <- gam(Num.Whelks ~ Year + Site, data = wd, method = "ML") # won't run with smoothed year
wgam1_summary <- summary(wgam1)
wgam1_summary$p.table
wgam1_summary$s.table
par(mfrow = c(1, 2))
plot(wgam1, all.terms = TRUE)
acf(residuals(wgam1)) 
```

```{r}
## Compare models - "Generally, the smaller the AIC, the “better” is the predictive performance of the model."
# Here we compare Linear model and GAM

AIC(whelks.m2, wgam1) # lower AIC = better performing model --> equal
#so we're using the linear model
```


##### Plot with Model Data

######Fig2D - Whelks Model Data
```{r}
Fig2D <- wd %>% 
  ggplot(aes(Year, Num.Whelks, color = Site, fill = Site)) + 
  geom_smooth(method = "lm", aes(group=NULL), fill = "gray79", color = "gray25") + 
  geom_point(size = 5, alpha = 0.7, pch = 21, color = "black") +
  theme_minimal() +
  theme(panel.background = element_blank(), 
        axis.line = element_line (colour = "black"), 
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.text = element_text(size = 25),
        axis.title = element_text(size = 25),
        legend.position = "none") + 
  labs(x = "Year", y = "Whelk Abundance") +
  scale_x_continuous(breaks = c(2014, 2016, 2018, 2020, 2022, 2024)) +
  scale_color_manual(values=group.colors) +
  scale_fill_manual(values=group.colors)

Fig2D
#ggsave("Fig2D.png", height=3, width=3, scale=3)

```