---
title: "Models"
author: "Micaela Chapuis"
date: "2024-07-16"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(broom)
```

```{r}
mussel_cover_yearly_avg <- read.csv(here("Data", "Data for Models", "mussel_cover_yearly_avg.csv"))
mussel_height_yearly_avg <- read.csv(here("Data", "Data for Models", "mussel_height_yearly_avg.csv"))

mussels <- read.csv(here("Data", "Data for Models", "mussels.csv"))
mussels_B <- read.csv(here("Data", "Data for Models", "mussels_B.csv"))

seastars_avg <- read.csv(here("Data", "Data for Models", "seastars_avg.csv"))
whelks_avg <- read.csv(here("Data", "Data for Models", "whelks_avg.csv"))

wavedata <- read.csv(here("Data", "wavedata.csv"))
```



## Biotic Factors
### Mussel Cover + Predators

Joining the data sets
```{r}
mussel_cover_predators <- left_join(mussel_cover_yearly_avg, seastars_avg, by=c("Year", "Site"))
mussel_cover_predators <- left_join(mussel_cover_predators, whelks_avg, by=c("Year", "Site"))
```

Centering and standardizing predictors. This function will standardize the data by default unless indicated with "scale = FALSE"
```{r}
mussel_cover_predators$whelks.z <- as.numeric(scale(mussel_cover_predators$Num.Whelks))
mussel_cover_predators$seastars.z <- as.numeric(scale(mussel_cover_predators$Num.Seastars))
mussel_cover_predators$year.z <- as.numeric(scale(mussel_cover_predators$Year))
```

###### Model
```{r}
cover.pred.model <- lm(PCover ~ year.z +Site + whelks.z + seastars.z, data = mussel_cover_predators)

summary(cover.pred.model)
tidy(cover.pred.model)
```

```{r}
plot(cover.pred.model)
qqnorm(resid(cover.pred.model))
qqline(resid(cover.pred.model))
hist(resid(cover.pred.model))

sjPlot::plot_model(cover.pred.model, type="diag")

tidy(cover.pred.model)
```


### Mussel Height + Predators

```{r}
mussel_height_predators <- left_join(mussel_height_yearly_avg, seastars_avg, by=c("Year", "Site"))
mussel_height_predators <- left_join(mussel_height_predators, whelks_avg, by=c("Year", "Site"))
```

Centering and standardizing predictors. This function will standardize the data by default unless indicated with "scale = FALSE"
```{r}
mussel_height_predators$whelks.z <- as.numeric(scale(mussel_height_predators$Num.Whelks))
mussel_height_predators$seastars.z <- as.numeric(scale(mussel_height_predators$Num.Seastars))
mussel_height_predators$year.z <- as.numeric(scale(mussel_height_predators$Year))
```

<!-- ###### Model -->
```{r}
height.pred.model <- lm(Avg.Height ~ year.z + whelks.z + seastars.z, data = mussel_height_predators)

summary(height.pred.model)
tidy(height.pred.model)
```


```{r}
plot(height.pred.model) 
qqnorm(resid(height.pred.model))
qqline(resid(height.pred.model))
hist(resid(height.pred.model))

sjPlot::plot_model(height.pred.model, type="diag")

```


## Abiotic Factors
```{r}
wavedata$Date <- as.Date(wavedata$Date)
wavedata$Year <- year(wavedata$Date)
```

Rename WVHT and WTMP
```{r}
wavedata <- wavedata %>% rename("Wave.Height" = "WVHT",
                                "Water.Temp" = "WTMP")
```


Function to calculate average of any variable (here WVHT and WTMP) for a given interval (here we want to take the date of mussel data collection and calculate the average of the variable for the 3 months before that day)

```{r}

calculate_average_variable <- function(end_date, data, variable) {
  
  # Convert the end date to date format
  end_date <- as.Date(end_date)
  
  # Find the start date of the interval by subtracting 3 months from the end date
  start_date <- as.Date(end_date - months(3))
  
  # Create sequence of dates between start and end dates
  date_sequence <- seq(from = start_date, to = end_date, by = "days")
  
  # Subset data to only include rows for dates in the interval
  data_subset <- data[data$Date %in% date_sequence,]
  
  # Calculate average of the variable for the interval
  average_variable <- mean(data_subset[[variable]])
  
  # Return the average of the variable
  return(average_variable)
}
```


<!-- #### Mussel Cover -->

Make new df "mussels.abiotic" with the same data as mussels + will add wave data

```{r}
mussels.abiotic <- mussels
```

Loop through mussel data (mussels) and calculate average Wave.Height for each date + add it to mussels.abiotic

```{r}
for (i in 1:nrow(mussels)) {
  # Get end date from mussels
  end_date <- mussels[i, "Date"]
  # Calculate average wave height for interval
  average_waveheight <- calculate_average_variable(end_date, wavedata, "Wave.Height")
  # Store result in mussels
  mussels.abiotic[i, "Wave.Height"] <- average_waveheight
}
```

Repeat with Water.Temp

```{r}
for (i in 1:nrow(mussels)) {
  # Get end date from mussels
  end_date <- mussels[i, "Date"]
  # Calculate average water temperature for interval
  average_watertemp <- calculate_average_variable(end_date, wavedata, "Water.Temp")
  # Store result in mussels
  mussels.abiotic[i, "Water.Temp"] <- average_watertemp
}
```

```{r}
mussels.abiotic <- mussels.abiotic %>% group_by(Date, Date.num, Wave.Height, Water.Temp) %>% summarize(PCover = mean(PCover))
```

Making csv with the averaged wave data to make Figure 2 with the model data
```{r}
#write.csv(mussels.abiotic, file = here("Data", "Data for Models", "mussels.abiotic.csv"), row.names = FALSE)  #to use in NOAA_Wave_Data script
```


Centering and standardizing predictors. This function will standardize the data by default unless indicated with "scale = FALSE"
```{r}
mussels.abiotic$wvht.z <- as.numeric(scale(mussels.abiotic$Wave.Height))
mussels.abiotic$wtmp.z <- as.numeric(scale(mussels.abiotic$Water.Temp))
mussels.abiotic$date.z <- as.numeric(scale(mussels.abiotic$Date.num))
```


###### Model
```{r}
cover.abiotic.model <- lm(PCover ~ date.z + wvht.z + wtmp.z, data = mussels.abiotic)

summary(cover.abiotic.model)
tidy(cover.abiotic.model)
```


```{r}
plot(cover.abiotic.model)
qqnorm(resid(cover.abiotic.model))
qqline(resid(cover.abiotic.model))
hist(resid(cover.abiotic.model))

sjPlot::plot_model(cover.abiotic.model, type="diag")
```

#### Mussel Bed Height

Make new df "mussels.abiotic.B" with the same data as mussels_B + will add wave data

```{r}
mussels.abiotic.B <- mussels_B
```

Loop through mussel data (mussels) and calculate average Wave.Height for each date + add it to mussels.abiotic

```{r}
for (i in 1:nrow(mussels_B)) {
  # Get end date from mussels_B
  end_date <- mussels_B[i, "Date"]
  # Calculate average wave height for interval
  average_waveheight <- calculate_average_variable(end_date, wavedata, "Wave.Height")
  # Store result in mussels
  mussels.abiotic.B[i, "Wave.Height"] <- average_waveheight
}
```

Repeat with Water.Temp

```{r}
for (i in 1:nrow(mussels_B)) {
  # Get end date from mussels
  end_date <- mussels_B[i, "Date"]
  # Calculate average water temperature for interval
  average_watertemp <- calculate_average_variable(end_date, wavedata, "Water.Temp")
  # Store result in mussels
  mussels.abiotic.B[i, "Water.Temp"] <- average_watertemp
}
```

```{r}
mussels.abiotic.B <- mussels.abiotic.B %>% group_by(Date, Date.num, Wave.Height, Water.Temp) %>% summarize(Avg.Height = mean(Avg.Height))
```

Centering and standardizing predictors. This function will standardize the data by default unless indicated with "scale = FALSE"
```{r}
mussels.abiotic.B$wvht.z <- as.numeric(scale(mussels.abiotic.B$Wave.Height))
mussels.abiotic.B$wtmp.z <- as.numeric(scale(mussels.abiotic.B$Water.Temp))
mussels.abiotic.B$date.z <- as.numeric(scale(mussels.abiotic.B$Date.num))
```

###### Model
```{r}
height.abiotic.model <- lm(Avg.Height ~ date.z + wvht.z + wtmp.z, data = mussels.abiotic.B)

summary(height.abiotic.model)
tidy(height.abiotic.model)
```


```{r}
plot(height.abiotic.model)
qqnorm(resid(height.abiotic.model))
qqline(resid(height.abiotic.model))
hist(resid(height.abiotic.model))

sjPlot::plot_model(height.abiotic.model, type="diag")
```



