---
title: "Mussel_Recruitment"
author: "Micaela Chapuis"
date: "2024-07-10"
output: html_document
---

```{r}
library("dplyr")
library("lubridate")
library("ggplot2")
```

Load in Recruitment data
```{r}
recruitment <- read.csv("Data/mussel_recruitment_data.csv")
```

Get rid of extra columns
```{r}
extra.columns <- names(recruitment) %in% c("Date.Processed", "Date.Sorted", "Initials", "X..of.Bivalves")

recruitment <- recruitment[,!extra.columns]
```

Changing column names
```{r}
names(recruitment)[1] <- "Date"
names(recruitment)[3] <- "WholePlot"
names(recruitment)[5] <- "Num.Recruits"
```

Turning Date into Date format (month-day-year)
```{r}
recruitment$Date <- as.Date(recruitment$Date, "%m/%d/%y")
```

Making Year column + making it into a factor
```{r}
recruitment[,"Year"] <- NA
recruitment$Year <- as.factor(format(recruitment$Date, "%Y"))
```

Add in Month column
```{r}
recruitment[, "Month"] <- NA
recruitment$Month <- as.factor(format(recruitment$Date, "%m"))
```

Changing Tuffy (tuffy number), WholePlot and Site to factors
```{r}
recruitment$Tuffy <- as.factor(recruitment$Tuffy)
recruitment$WholePlot <- as.factor(recruitment$WholePlot)
recruitment$Site <- as.factor(recruitment$Site)
```

#### Fig 4 - Mussel Recruits by Year

Log transforming data (+1 because we have some 0s and their log would be infinity)
```{r}
recruitment[, "log_recruits"] <- NA
recruitment$log_recruits <- log(recruitment$Num.Recruits + 1)
```

```{r}
Fig4 <- recruitment %>% group_by(Year) %>% 
  ggplot(aes(x = Year, y = log_recruits)) + 
  geom_boxplot() +
  theme(panel.background = element_blank(),     
        axis.line = element_line(colour = "black"),
        axis.text = element_text(size=28), 
        axis.title = element_text(size=34)) +
  labs(x = "Year", y ="Number of Mussel Recruits (log (x+1))") + 
  geom_point(alpha=0.5) + geom_jitter(width=0.2, height =0.1)

Fig4
#ggsave("Fig4.png", height=2, width=3,scale = 3)
```
#### Recruitment ANOVA

```{r}
recruit_log <- recruitment %>% filter(!is.na(Num.Recruits)) %>% mutate(logRecruit = log(Num.Recruits+1))
```

https://broom.tidymodels.org/reference/tidy.anova.html
```{r}
recruit_anova <- aov(logRecruit ~ Year, data = recruit_log)
summary(recruit_anova)
tidy(recruit_anova)
```

```{r}
#checking normality
par(mfrow = c(1,2)) # puts two plots side by side in the same window
hist(recruit_anova$residuals) # residuals histogram
plot(recruit_anova, which = 2) # Q-Q plot
```


```{r}
#checking homoskedasticity (homogeneity of variances) w/ residuals vs fitted plot
plot(recruit_anova, which = 1)
# We want to see a straight red line centered around zero! This means residuals do NOT systematically differ across different groups. --> looks pretty good
```

#### Recruitment lm

```{r}
recruit_lm <- lm(logRecruit ~ Year, data = recruit_log)
summary(recruit_lm)
```


#### TukeyHSD

```{r}
recruit_tukey <- TukeyHSD(recruit_anova, conf.level = .95)
recruit_tukey
plot(recruit_tukey)
tidy(recruit_tukey)
```